---
title: "Linear regression"
output:
  bookdown::html_document2:
    #base_format: rmarkdown::html_vignette
    #highlight: tango
    number_sections: true
    toc: true
    toc_float: true
    fig_caption: yes
link-citations: yes
bibliography: bib/biblio-jSDM.bib
biblio-style: bib/jae.bst
csl: bib/journal-of-applied-ecology.csl
pkgdown:
  as_is: true
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteIndexEntry{Linear regression}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  fig.retina=1,
  fig.width = 6, fig.height = 6,
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  highlight = TRUE
)
```

# Model definition 

We define the following model :

- Response variable: $Y=(y_{ij})^{i=1,\ldots,nsite}_{j=1,\ldots,nsp}$ such that $y_{ij} \sim \mathcal{N}(\theta_{ij}, V)$ such that: 

$$ \theta_{ij} =\alpha_i + \beta_{0j}+X_i.\beta_j+ W_i.\lambda_j $$




It can be easily shown that: $y_{ij} = \alpha_i + \beta_{0j} + X_i.\beta_j + W_i.\lambda_j + \epsilon_{i,j}$, with $\forall (i,j) \ \epsilon_{ij} \sim \mathcal{N}(0,V)$ where:

- Latent variables: $W_i=(W_i^1,\ldots,W_i^q)$ where $q$ is the number of latent variables considered, which has to be fixed by the user (by default $q=2$).
We assume that $W_i \sim \mathcal{N}(0,I_q)$ and we define the associated coefficients: $\lambda_j=(\lambda_j^1,\ldots, \lambda_j^q)'$. We use a prior distribution $\mathcal{N}(0,1)$ for all lambdas not concerned by constraints to $0$ on upper diagonal and to strictly positive values on diagonal. 

- Explanatory variables: bioclimatic data about each site. $X=(X_i)_{i=1,\ldots,nsite}$ with $X_i=(x_i^1,\ldots,x_i^p)\in \mathbb{R}^p$ where $p$ is the number of bioclimatic variables considered.
The corresponding regression coefficients for each species $j$ are noted : $\beta_j=(\beta_j^1,\ldots,\beta_j^p)'$.

- $\beta_{0j}$ correspond to the intercept for species $j$ which is assumed to be a fixed effect. We use a prior distribution $\mathcal{N}(0,1)$ for all betas. 

- $\alpha_i$ represents the random effect of site $i$ such as $\alpha_i \sim \mathcal{N}(0,V_{\alpha})$ and we assume that $V_{\alpha} \sim \mathcal {IG}(\text{shape}=0.5, \text{rate}=0.005)$ as prior distribution by default. 

- $V$ represents the variance of residuals or overdispersion term and we assume that $V \sim \mathcal {IG}(\text{shape}=0.5, \text{rate}=0.005)$ as prior distribution by default. 


# Simulated explanatory variables 

We simulate two explanatory variables considering 150 inventory sites. 

```{r env}
#== Data simulation

#= Number of inventory sites
nsite <- 150

#= Set seed for repeatability
seed <- 1234
set.seed(seed)

#= Ecological process (suitability)
x1 <- rnorm(nsite, 0, 1)
x2 <- rnorm(nsite, 0, 1)
X <- cbind(rep(1, nsite), x1, x2)
colnames(X) <- c("Intercept", "Covariate_1", "Covariate_2")
np <- ncol(X)

head(X)
```


# Simulated response data-set

We simulate the response variable for 20 species on the 150 inventory sites by considering the explanatory variables simulated above and following the model defined previously with two latent axes.

```{r response-data}
#= Number of species
nsp <- 20

#= Number of latent variables
n_latent <- 2

#= Latent variables W
W <- matrix(rnorm(nsite*n_latent,0,1), nsite, n_latent)
#= Fixed species effect beta 
beta.target <- t(matrix(runif(nsp*np,-1,1),
                        byrow=TRUE, nrow=nsp))
#= Factor loading lambda  
lambda.target <- matrix(0, n_latent, nsp)
mat <- t(matrix(runif(nsp*n_latent, -1, 1), byrow=TRUE, nrow=nsp))
lambda.target[upper.tri(mat, diag=TRUE)] <- mat[upper.tri(mat, diag=TRUE)]
diag(lambda.target) <- runif(n_latent, 0, 2)
#= Variance of random site effect 
V_alpha.target <- 0.2
#= Random site effect alpha
alpha.target <- rnorm(nsite,0 , sqrt(V_alpha.target))
# Simulation of response data with probit link
theta.target <- X%*%beta.target + W%*%lambda.target + alpha.target
V.target <- 0.2
Y <- matrix(rnorm(nsite*nsp, theta.target, sqrt(V.target)), nrow=nsite)
colnames(Y) <- paste0("sp_", 1:nsp)
rownames(Y) <- paste0("site_", 1:nsite)

head(Y)
```


This data-set is available in the [`jSDM-package`](https://ecology.ghislainv.fr/jSDM/reference/jSDM-package.html) R package. It can be loaded with the `data()` command. The [`frogs`](https://ecology.ghislainv.fr/jSDM/reference/frogs.html) dataset is in "wide" format: each line is a site and the occurrence data (from Species_1 to Species_9) are in columns. A site is characterized by its x-y geographical coordinates, one discrete covariate and two other continuous covariates. 

We rearrange the data in two data-sets: a first one for the presence-absence observations for each species (columns) at each site (rows), and a second one for the site characteristics.

We also normalize the continuous explanatory variables to facilitate MCMC convergence. 

# Parameter inference

We use the `jSDM_gaussian` function to fit the JSDM (increase the number of iterations to achieve convergence). 

```{r jSDM-probit}
library(jSDM)

# Increase number of iterations (burnin and mcmc) to get convergence
mod <- jSDM_gaussian(# Iteration
  burnin=1000,
  mcmc=1000,
  thin=1,
  # Response variable
  response_data=Y,
  # Explanatory variables
  site_formula =~ Covariate_1 + Covariate_2,
  site_data=X,
  n_latent=2,
  site_effect="random",
  # Starting values
  alpha_start=0,
  beta_start=0,
  lambda_start=0,
  W_start=0,
  V_alpha=1,
  V_start=1 ,
  # Priors
  shape_Valpha=0.5,
  rate_Valpha=0.0005,
  shape_V=0.5, rate_V=0.0005,
  mu_beta=0, V_beta=1,
  mu_lambda=0, V_lambda=1,
  seed=1234, verbose=1)
```

# Analysis of the results

```{r plot-results-probit}
np <- nrow(mod$model_spec$beta_start)

# ===================================================
# Result analysis
# ===================================================

#==========
#== Outputs

#= Parameter estimates

# Species effects beta and factor loadings lambda

## Trace and density of beta_j for the first two species
mean_beta <- matrix(0,nsp,ncol(X))
par(mfrow=c(2,2))
for (j in 1:nsp){
  mean_beta[j,] <- apply(mod$mcmc.sp[[j]]
                         [,1:ncol(X)], 2, mean) 
  if(j<=2){
    for (p in 1:ncol(X)){
      coda::traceplot(mod$mcmc.sp[[j]][,p])
      coda::densplot(mod$mcmc.sp[[j]][,p],
                     main = paste(colnames(mod$mcmc.sp[[j]])[p],", species : ",j))
      abline(v=beta.target[p,j],col='red')
    }
  }
}

## Trace and density of lambda_j for the first two species
mean_lambda <- matrix(0,nsp,n_latent)
par(mfrow=c(2,2))
for (j in 1:nsp){
  mean_lambda[j,] <- apply(mod$mcmc.sp[[j]]
                           [,(ncol(X)+1):(ncol(X)+n_latent)], 2, mean)  
  if(j<=2){
    for (l in 1:n_latent) {
      coda::traceplot(mod$mcmc.sp[[j]][,ncol(X)+l])
      coda::densplot(mod$mcmc.sp[[j]][,ncol(X)+l],
                     main=paste(colnames(mod$mcmc.sp[[j]])
                                [ncol(X)+l],", species : ",j))
      abline(v=lambda.target[l,j],col='red')
    }
  }
}
# Representation of fitted values according to expected ones
par(mfrow=c(1,2))
plot(t(beta.target), mean_beta,
     main="species effect beta",
     xlab ="obs", ylab ="fitted")
abline(a=0,b=1,col='red')
plot(t(lambda.target), mean_lambda,
     main="factor loadings lambda",
     xlab ="obs", ylab ="fitted")
abline(a=0,b=1,col='red')

## W latent variables
## Trace and density of latent variables W_i for the first two sites
par(mfrow=c(2,2))
for (l in 1:n_latent) {
  for (i in 1:2) {
    coda::traceplot(mod$mcmc.latent[[paste0("lv_",l)]][,i],
                    main = paste0("Latent variable W_", l, ", site ", i))
    coda::densplot(mod$mcmc.latent[[paste0("lv_",l)]][,i],
                   main = paste0("Latent variable W_", l, ", site ", i))
    abline(v=W[i,l],col='red')
  }
}
# Representation of fitted values according to expected ones
par(mfrow=c(1,2))
for (l in 1:n_latent) {
  plot(W[,l],
       summary(mod$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"],
       main = paste0("Latent variable W_", l),
       xlab ="obs", ylab ="fitted")
  abline(a=0,b=1,col='red')
}

## alpha
## Trace and density of alpha_i for the first two sites
for (i in 1:2){
  coda::traceplot(mod$mcmc.alpha[,i],
                  main = paste0("Site effect alpha_", i))
  coda::densplot(mod$mcmc.alpha[,i], 
                 main = paste0("Site effect alpha_", i))
  abline(v=alpha.target[i],col='red')
}
# Representation of fitted values according to expected ones
par(mfrow=c(1,3))
plot(alpha.target, summary(mod$mcmc.alpha)[[1]][,"Mean"],
     xlab ="obs", ylab ="fitted", main="site effect alpha")
abline(a=0,b=1,col='red')

## Valpha
coda::traceplot(mod$mcmc.V_alpha)
coda::densplot(mod$mcmc.V_alpha)
abline(v=V_alpha.target,col='red')

## Variance of residuals
par(mfrow=c(1,2))
coda::traceplot(mod$mcmc.V)
coda::densplot(mod$mcmc.V,
               main="Variance of residuals")
abline(v=V.target, col='red')

## Deviance
summary(mod$mcmc.Deviance)
plot(mod$mcmc.Deviance)

#= Predictions
par(mfrow=c(1,1))
plot(Y, mod$Y_pred,
     main="Response variable",
     xlab="obs", ylab="fitted")
abline(a=0,b=1,col='red')
```

# Matrice of correlations 

After fitting the jSDM with latent variables, the **full species residual correlation matrix** $R=(R_{ij})^{i=1,\ldots, nspecies}_{j=1,\ldots, nspecies}$ can be derived from the covariance in the latent variables such as : 
$$\Sigma_{ij} = 
\begin{cases} 
\lambda_i^T .\lambda_j + V, & \text{if } i=j \\
\lambda_i^T .\lambda_j, & \text{else.}
\end{cases}$$, then we compute correlations from covariances :
$$R_{i,j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma _{ii}\Sigma _{jj}}}$$.

We use the `plot_residual_cor()` function to compute and display  the residual correlation matrix :

```{r correlation-matrix-probit}
plot_residual_cor(mod)
```

# Predictions 

We use the `predict.jSDM()` S3 method on the `mod` object of class `jSDM` to compute the mean (or expectation) of the posterior distributions obtained and get the expected values of response variable.

```{r predictions-probit}
# Sites and species concerned by predictions :
## 50 sites among the 150
Id_sites <- sample.int(150, 50)
## All species 
Id_species <- colnames(Y)
# Simulate new observations of covariates on those sites 
simdata <- matrix(nrow=50, ncol = ncol(mod$model_spec$site_data))
colnames(simdata) <- colnames(mod$model_spec$site_data)
rownames(simdata) <- Id_sites
simdata <- as.data.frame(simdata)
simdata$Covariate_1 <- rnorm(50)
simdata$Covariate_2 <- rnorm(50)

# Predictions 
Y_pred <- predict(mod, newdata=simdata, Id_species=Id_species,
                      Id_sites=Id_sites, type="mean")
hist(Y_pred, main="Predicted response with simulated data", xlab="Y_pred")
```

# References