---
title: "Comparison jSDM-boral"
output:
  bookdown::html_document2:
    #base_format: rmarkdown::html_vignette
    #highlight: tango
    number_sections: true
    toc: true
    #toc_float: true
    fig_caption: yes
link-citations: yes
bibliography: bib/biblio-jSDM.bib
biblio-style: bib/jae.bst
csl: bib/journal-of-applied-ecology.csl
pkgdown:
  as_is: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Comparison jSDM-boral}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align="center",
  fig.width=6, fig.height=6,
  cache=FALSE,
  collapse=TRUE,
  comment="#>",
  highlight=TRUE,
  eval=FALSE
)
```

# Package boral

## Simulated dataset

300 sites et 100 species with as parameters :

```{r simulated-data, echo=TRUE}

# Toy example : 
nsp <- 100
nsite <- 300
np <- 3
nl <- 2
seed <- 123
set.seed(seed)

# Ecological process (suitability)
x1 <- rnorm(nsite,0,1)
x2 <- rnorm(nsite,0,1)
X <- cbind(rep(1,nsite),x1,x2)
# Simulation of parameters
W <- cbind(rnorm(nsite,0,1),rnorm(nsite,0,1))
data=cbind (X,W)
beta.target <- t(matrix(runif(nsp*np,-2,2), byrow=TRUE, nrow=nsp))
l.zero <- 0
l.diag <- runif(2,0,2)
l.other <- runif(nsp*2-3,-2,2)
lambda.target <- t(matrix(c(l.diag[1],l.zero,l.other[1],l.diag[2],l.other[-1]), byrow=T, nrow=nsp))
param.target <- rbind(beta.target,lambda.target)
param.target=rbind(beta.target,lambda.target)
V_alpha.target=0.5
V=1
alpha.target <- rnorm(nsite,0,sqrt(V_alpha.target))

# Simulation of response data with probit link
probit_theta=X %*% beta.target + W %*% lambda.target + alpha.target
e=matrix(rnorm(nsp*nsite,0,sqrt(V)),nsite,nsp)
Z_true <- probit_theta + e
Y=matrix (NA, nsite,nsp)
for (i in 1:nsite){
  for (j in 1:nsp){
    if ( Z_true[i,j] > 0) {Y[i,j] <- 1}
    else {Y[i,j] <- 0}
  }
}
```

- $\beta_{target}$:

```{r beta}
library(knitr)
colnames(beta.target) <- paste(rep("nsp",nsp),1:nsp,sep="")
kable(beta.target[,1:15], digits=1)
```

- $\lambda_{target}$:

```{r lambda}
colnames(lambda.target) <- paste(rep("nsp",nsp),1:nsp,sep="")
kable(lambda.target[,1:15], digits=1)
```

- $V_{\alpha}$: 

```{r V_alpha}
V_alpha.target
```

```{r boral-simulation, cache=TRUE}
library(boral)
T1<-Sys.time() 
mod_boral_sim <- boral(y=Y, X=X[,-1], lv.control=list(num.lv=nl, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nsite),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=10000, n.iteration=15000,n.thin=5,seed=123))
T2<-Sys.time() 
T_boral_sim=difftime(T2, T1)

# Predicted probit theta 
probit_theta_pred_sim=mod_boral_sim$row.coefs[[1]]$mean + X[,-1] %*% t(mod_boral_sim$X.coefs.mean)  + mod_boral_sim$lv.coefs.mean[,"beta0"] + mod_boral_sim$lv.mean%*%t(mod_boral_sim$lv.coefs.mean[,-1])
# RMSE
SE=(probit_theta-probit_theta_pred_sim)^2
RMSE_boral_sim=sqrt(sum(SE/(nsite*nsp)))
# Deviance 
logL=0
for (i in 1:nsite){
  for (j in 1:nsp){
    theta <- pnorm(probit_theta_pred_sim[i,j])
    logL=logL + dbinom(Y[i,j],1,theta,1)  
  }
}
Deviance_boral_sim <- -2*logL
```

```{r boral-simulation-plot, cache=T}
## beta_j
par(mfrow=c(1,2))
plot(t(beta.target),cbind(mod_boral_sim$lv.coefs.mean[,1], mod_boral_sim$X.coefs.mean), xlab="betas target", ylab="means of estimated betas") 
abline(a=0,b=1,col='red')
## lambda_j
plot(t(lambda.target),mod_boral_sim$lv.coefs.mean[,-1], xlab="lambdas target", ylab="means of estimated lambdas") 
abline(a=0,b=1,col='red')
## Latent variable W 
par(mfrow=c(1,2))
for (l in 1:nl) {
  plot(W[,l],mod_boral_sim$lv.mean[,l], main=paste0("Latent variable W_", l), xlab="W_i target", ylab="means of estimated W_i")
  abline(a=0,b=1,col='red')
}
## alphas 
plot(alpha.target, mod_boral_sim$row.coefs[[1]]$mean, xlab="alphas target", ylab="means of estimated alphas")
abline(a=0,b=1,col='red')
print(paste("Estimated V_alpha : ",mod_boral_sim$row.sigma[[1]]$mean))
## Prediction
# probit_theta_pred 
plot(probit_theta,probit_theta_pred_sim, main="Prediction", xlab ="True probit theta", ylab=" Estimated probit theta")
abline(a=0,b=1,col='red')
```

## Mosquito dataset

```{r boral-mosquito, cache=TRUE}
# Import center and reduce Mosquito dataset
data(mosquitos, package="jSDM")
head(mosquitos)
Env_mosquito <- mosquitos[,17:29]
Env_mosquito <- cbind(scale(Env_mosquito[,1:4]), Env_mosquito[,5:13])
PA_mosquito <- mosquitos[,1:16]

# Fit the model 
T1 <- Sys.time()
mod_boral_Mosquito <- boral(y=PA_mosquito, X=Env_mosquito, lv.control=list(num.lv=2, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nrow(PA_mosquito)),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=10000, n.iteration=15000,n.thin=5,seed=123))
T2 <- Sys.time()
T_boral_Mosquito <- difftime(T2,T1)

# Predicted probit theta 
probit_theta_pred_Mosquito=mod_boral_Mosquito$row.coefs[[1]]$mean + as.matrix(Env_mosquito) %*% t(mod_boral_Mosquito$X.coefs.mean)  + mod_boral_Mosquito$lv.coefs.mean[,"beta0"] + mod_boral_Mosquito$lv.mean%*%t(mod_boral_Mosquito$lv.coefs.mean[,-1])

# Deviance
logL=0
for (i in 1:nrow(PA_mosquito)){
  for (j in 1:ncol(PA_mosquito)){
    theta <- pnorm(probit_theta_pred_Mosquito[i,j])
    logL=logL + dbinom(PA_mosquito[i,j],1,theta,1)  
  }
}
Deviance_boral_Mosquito <- -2*logL
```


## Eucalypts dataset

```{r boral-eucalypts, cache=TRUE}
# Import center and reduce Eucalypts dataset
data(eucalypts, package="jSDM")
head(eucalypts)
Env_Eucalypts <- cbind(scale(eucalypts[,c("Rockiness","VallyBotFlat","PPTann", "cvTemp","T0")]),eucalypts[,c("Sandiness","Loaminess")])
PA_Eucalypts <- eucalypts[,1:12]

# Fit the model 
T1 <- Sys.time()
mod_boral_Eucalypts <- boral(y=PA_Eucalypts, X=Env_Eucalypts, lv.control=list(num.lv=2, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nrow(PA_Eucalypts)),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=10000, n.iteration=15000,n.thin=5,seed=123))
T2 <- Sys.time()
T_boral_Eucalypts <- difftime(T2,T1)

# Predicted probit theta 
probit_theta_pred_Eucalypts=mod_boral_Eucalypts$row.coefs[[1]]$mean + as.matrix(Env_Eucalypts) %*% t(mod_boral_Eucalypts$X.coefs.mean)  + mod_boral_Eucalypts$lv.coefs.mean[,"beta0"] + mod_boral_Eucalypts$lv.mean%*%t(mod_boral_Eucalypts$lv.coefs.mean[,-1])

# Deviance
logL=0
for (i in 1:nrow(PA_Eucalypts)){
  for (j in 1:ncol(PA_Eucalypts)){
    theta <- pnorm(probit_theta_pred_Eucalypts[i,j])
    logL=logL + dbinom(PA_Eucalypts[i,j],1,theta,1)  
  }
}
Deviance_boral_Eucalypts <- -2*logL
```

## Frogs dataset

```{r boral-frogs, cache=TRUE}
# Import center and reduce Frogs dataset
data(frogs, package="jSDM")
head(frogs)
Env_Frogs <- cbind(scale(frogs[,"Covariate_1"]),frogs[,"Covariate_2"],scale(frogs[,"Covariate_3"]))
colnames(Env_Frogs) <- colnames(frogs[,1:3])
PA_Frogs <- frogs[,4:12]

# Fit the model
T1 <- Sys.time()
mod_boral_Frogs <- boral(y=PA_Frogs, X=Env_Frogs, lv.control=list(num.lv=2, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nrow(PA_Frogs)),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=10000, n.iteration=15000,n.thin=5,seed=123))
T2 <- Sys.time()
T_boral_Frogs <- difftime(T2,T1)

# Predicted probit theta 
probit_theta_pred_Frogs=mod_boral_Frogs$row.coefs[[1]]$mean + as.matrix(Env_Frogs) %*% t(mod_boral_Frogs$X.coefs.mean)  + mod_boral_Frogs$lv.coefs.mean[,"beta0"] + mod_boral_Frogs$lv.mean%*%t(mod_boral_Frogs$lv.coefs.mean[,-1])

# Deviance
logL=0
for (i in 1:nrow(PA_Frogs)){
  for (j in 1:ncol(PA_Frogs)){
    theta <- pnorm(probit_theta_pred_Frogs[i,j])
    logL=logL + dbinom(PA_Frogs[i,j],1,theta,1)  
  }
}
Deviance_boral_Frogs <- -2*logL
```

## Fungi dataset

```{r boral-fungi, cache=TRUE}

# Import center and reduce fungi dataset
data(fungi, package="jSDM")
Env_Fungi <- cbind(scale(fungi[,c("diam","epi","bark")]),fungi[,c("dc1","dc2","dc3","dc4","dc5","quality3","quality4","ground3","ground4")])
colnames(Env_Fungi) <- c("diam","epi","bark","dc1","dc2","dc3","dc4","dc5","quality3","quality4","ground3","ground4")
PA_Fungi <- fungi[,c("antser","antsin","astfer","fompin","hetpar","junlut","phefer","phenig","phevit","poscae","triabi")]

# Fit the model 
T1 <- Sys.time()
mod_boral_Fungi <- boral(y=PA_Fungi, X=Env_Fungi, lv.control=list(num.lv=2, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nrow(PA_Fungi)),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=10000, n.iteration=15000,n.thin=5,seed=123))
T2 <- Sys.time()
T_boral_Fungi <- difftime(T2,T1)

# Predicted probit theta 
probit_theta_pred_Fungi=mod_boral_Fungi$row.coefs[[1]]$mean + as.matrix(Env_Fungi) %*% t(mod_boral_Fungi$X.coefs.mean)  + mod_boral_Fungi$lv.coefs.mean[,"beta0"] + mod_boral_Fungi$lv.mean%*%t(mod_boral_Fungi$lv.coefs.mean[,-1])

# Deviance
logL=0
for (i in 1:nrow(PA_Fungi)){
  for (j in 1:ncol(PA_Fungi)){
    theta <- pnorm(probit_theta_pred_Fungi[i,j])
    logL=logL + dbinom(PA_Fungi[i,j],1,theta,1)  
  }
}
Deviance_boral_Fungi <- -2*logL
```

# Package jSDM

## Simulated dataset

```{r jSDM-simulation, cache=TRUE}
library(jSDM)

# Fit the model
T1<-Sys.time() 
mod_jSDM_block_sim <- jSDM_binomial_probit_block(
  # Chains
  burnin=10000, mcmc=5000, thin=5,
  # Response variable
  presence_site_sp=Y, 
  # Explanatory variables 
  site_suitability=~x1+x2,   
  site_data=X,
  # Model specification
  n_latent=2, site_effect="random",
  # Starting values
  alpha_start=0, beta_start=0,
  lambda_start=0, W_start=0,
  V_alpha=1, 
  # Priors 
  shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  # Various 
  seed=1234, verbose=1)
T2<-Sys.time() 
T_jSDM_block_sim=difftime(T2, T1)

# RMSE
SE=(probit_theta-mod_jSDM_block_sim$probit_theta_pred)^2
RMSE_jSDM_block_sim=sqrt(sum(SE/(nsite*nsp)))
```

```{r jSDM-simulation-plot}

# ===================================================
# Result analysis
# ===================================================

## V_alpha
par(mfrow=c(1,2))
coda::traceplot(mod_jSDM_block_sim$mcmc.V_alpha, main="V_alpha")
coda::densplot(mod_jSDM_block_sim$mcmc.V_alpha, main="V_alpha")
abline(v=V_alpha.target,col='red')

## Deviance
plot(mod_jSDM_block_sim$mcmc.Deviance, main="Deviance")

## beta_j
mean_betas <- matrix(0,nsp,np)
par(mfrow=c(ncol(X),2))
for (j in 1:nsp) {
  for (p in 1:ncol(X)) {
    mean_betas[j,p] <-mean(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][j,p])
    if (j < 3){
      coda::traceplot(coda::as.mcmc(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][,p]))
      coda::densplot(coda::as.mcmc(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][,p]), 
                     main=paste(colnames(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]])[p],", species : ",j))
      abline(v=beta.target[p,j],col='red')
    }
  }
}

## lambda_j
mean_lambdas <- matrix(0,nsp,nl)
par(mfrow=c(nl,2))
for (j in 1:nsp) {
  for (l in 1:nl) {
        mean_lambdas[j,l] <-mean(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][j,ncol(X)+l])

    if (j < 3){
      coda::traceplot(coda::as.mcmc(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][,ncol(X)+l]))
      coda::densplot(coda::as.mcmc(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][,ncol(X)+l]), 
                     main=paste(colnames(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]])[ncol(X)+l],", species : ",j))
      abline(v=lambda.target[l,j],col='red')
    }
  }
}
par(mfrow=c(1,2))
plot(t(beta.target),mean_betas, xlab="betas target", ylab="means of estimated betas")
abline(a=0,b=1,col='red')
plot(t(lambda.target),mean_lambdas, xlab="lambdas target", ylab="means of estimated lambda")
abline(a=0,b=1,col='red')

## W latent variables
par(mfrow=c(1,2))
for (l in 1:nl) {
  plot(W[,l],summary(mod_jSDM_block_sim$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"], main=paste0("Latent variable W_", l), xlab="W_i target", ylab="means of estimated W_i")
  abline(a=0,b=1,col='red')
}

## alpha
plot(alpha.target,summary(mod_jSDM_block_sim$mcmc.alpha)[[1]][,"Mean"], xlab= "alphas target", ylab="means of estimated alphas")
abline(a=0,b=1,col='red')

#= Predictions

## probit_theta
par(mfrow=c(1,2))
plot(probit_theta,mod_jSDM_block_sim$probit_theta_pred, xlab="True probit theta", ylab="Predicted probit theta")
abline(a=0,b=1,col='red')
title(main="Predictions")
## Z
plot(Z_true,mod_jSDM_block_sim$Z_latent, xlab="True latent variable Z", ylab="Predicted latent variable Z")
abline(a=0,b=1,col='red')
```

## Mosquito dataset

```{r jSDM-mosquito, cache=TRUE}
# Fit the model
T1 <- Sys.time()
mod_jSDM_block_Mosquito <- jSDM_binomial_probit_block(
  # Chains 
  burnin=10000, mcmc=5000, thin=5,
  # Response variable 
  presence_site_sp=PA_mosquito, 
  # Explanatory variables 
  site_suitability=~.,   
  site_data=Env_mosquito,
  # Model specification
  site_effect="random", n_latent=2,
  # Starting values 
  alpha_start=0, beta_start=0,
  lambda_start=0, W_start=0,
  V_alpha=1,
  # Priors 
  shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  # Various 
  seed=1234, verbose=1)
T2 <- Sys.time()
T_jSDM_block_Mosquito <- difftime(T2,T1)
```

## Eucalypts dataset

```{r jSDM-eucalypts, cache=TRUE}
# Fit the model
T1 <- Sys.time()
mod_jSDM_block_Eucalypts <- jSDM_binomial_probit_block(
  # Chains
  burnin=10000, mcmc=5000, thin=5,
  # Response variable 
  presence_site_sp=PA_Eucalypts,
  # Explanatory variables 
  site_suitability=~.,   
  site_data=Env_Eucalypts,
  # Model specification
  n_latent=2, site_effect="random",
  # Starting values 
  alpha_start=0, beta_start=0,
  lambda_start=0, W_start=0,
  V_alpha=1,
  # Priors 
  shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  # Various 
  seed=1234, verbose=1)
T2 <- Sys.time()
T_jSDM_block_Eucalypts <- difftime(T2,T1)
```

## Frogs dataset

```{r jSDM-frogs, cache=TRUE}
# Fit the model 
T1 <- Sys.time()
mod_jSDM_block_Frogs <- jSDM_binomial_probit_block(
  # Chains
  burnin=10000, mcmc=5000, thin=5,
  # Response variable 
  presence_site_sp=as.matrix(PA_Frogs), 
  # Explanatory variables 
  site_suitability=~.,   
  site_data=as.data.frame(Env_Frogs),
  # Model specification 
  n_latent=2, site_effect="random",
  # Starting values 
  alpha_start=0, beta_start=0,
  lambda_start=0, W_start=0,
  V_alpha=1,
  # Priors 
  shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  # Various 
  seed=1234, verbose=1)
T2 <- Sys.time()
T_jSDM_block_Frogs <- difftime(T2,T1)
```

## Fungi dataset

```{r jSDM-fungi, cache=TRUE}
# Fit the model
T1 <- Sys.time()
mod_jSDM_block_Fungi <- jSDM_binomial_probit_block(
  # Chains 
  burnin=10000, mcmc=5000, thin=5,
  # Response variable 
  presence_site_sp=PA_Fungi, 
  # Explanatory variables 
  site_suitability=~.,   
  site_data=Env_Fungi,
  # Model specification
  n_latent=2, site_effect="random",
  # Starting values 
  alpha_start=0, beta_start=0,
  lambda_start=0, W_start=0,
  V_alpha=1,
  # Priors 
  shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  # Various 
  seed=1234, verbose=1)
T2 <- Sys.time()
T_jSDM_block_Fungi <- difftime(T2,T1)
```

# Comparison

## Root-Mean-Square Error (RMSE) for simulated data

```{r RMSE}
library(knitr)
result <- data.frame(matrix(NA,1,2),row.names=  c("RMSE"))
colnames(result) <- c("Boral","jSDM_block")
result$Boral <- RMSE_boral_sim
result$jSDM_block <- RMSE_jSDM_block_sim
kable(result, digits =1)
```

## Compilation time in secondes

```{r time}
library(knitr)
result <- data.frame(matrix(NA,2,5),row.names=c("time_boral","time_jSDM_block"))
colnames(result) <- c("Simulated","Mosquito","Eucalypts","Frogs","Fungi")
result[1,]=c(T_boral_sim, T_boral_Mosquito, T_boral_Eucalypts, T_boral_Frogs, T_boral_Fungi)
result[2,]=c(T_jSDM_block_sim, T_jSDM_block_Mosquito, T_jSDM_block_Eucalypts, T_jSDM_block_Frogs, T_jSDM_block_Fungi)
kable(result, digits =1)
```

## Deviance

```{r deviance}
library(knitr)
result <- data.frame(matrix(NA,2,5),row.names=c("Deviance boral", "Deviance jSDM_block"))
colnames(result) <- c("Simulated","Mosquito","Eucalypts","Frogs","Fungi")
result[1,] <- c(Deviance_boral_sim, Deviance_boral_Mosquito, Deviance_boral_Eucalypts, Deviance_boral_Frogs, Deviance_boral_Fungi)
result[2,] <- c( mean(mod_jSDM_block_sim$mcmc.Deviance),  mean(mod_jSDM_block_Mosquito$mcmc.Deviance),  mean(mod_jSDM_block_Eucalypts$mcmc.Deviance),  mean(mod_jSDM_block_Frogs$mcmc.Deviance),  mean(mod_jSDM_block_Fungi$mcmc.Deviance))
kable(result, digits =1)
```

## Estimated  Parameters

### Simulated dataset

300 sites 100 species.

```{r jSDM-boral-simulation}
# Alpha et V_alpha

par(mfrow=c(1,2))
plot(mod_boral_sim$row.coefs[[1]]$mean,summary(mod_jSDM_block_sim$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by Boral", ylab="alphas estimated by jSDM", main="Site random effect alpha")
abline(a=0,b=1,col='red')
points(mod_boral_sim$row.sigma[[1]]$mean,summary(mod_jSDM_block_sim$mcmc.V_alpha)[[1]]["Mean"],
       pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)
# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
  for (p in 1:np){
    jSDM_betas[j,p] <- mean(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][,p])
  }
}
boral_betas <- cbind(mod_boral_sim$lv.coefs.mean[,"beta0"],mod_boral_sim$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by Boral", ylab="betas estimated by jSDM", main="Species fixed effect beta_j")
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
  for (l in 1:nl){
    jSDM_lambdas[j,l] <- mean(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][,np+l])
  }
}
boral_lambdas <- mod_boral_sim$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by Boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda_j")
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- colMeans(mod_jSDM_block_sim$mcmc.latent[[paste0("lv_",l)]])
}
plot(mod_boral_sim$lv.mean, jSDM_lvs, xlab="latent variables estimated by Boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2")
abline(a=0,b=1,col='red')

# Predictions 
plot(probit_theta_pred_sim, mod_jSDM_block_sim$probit_theta_pred, xlab="probit theta estimated by Boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta")
abline(a=0,b=1,col='red')
```

### Mosquito dataset

```{r jSDM-boral-mosquito}
print(paste(nrow(PA_mosquito),"sites and ",ncol(PA_mosquito)," species"))
nsp <- ncol(mod_jSDM_block_Mosquito$model_spec$presences)
nsite <- nrow(mod_jSDM_block_Mosquito$model_spec$presences)
nl <- mod_jSDM_block_Mosquito$model_spec$n_latent
np <- nrow(mod_jSDM_block_Mosquito$model_spec$beta_start)
# Alpha et V_alpha
par(mfrow=c(1,2))
plot(mod_boral_Mosquito$row.coefs[[1]]$mean,summary(mod_jSDM_block_Mosquito$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by Boral", ylab="alphas estimated by jSDM", main="Site random effect alpha")
abline(a=0,b=1,col='red')
points(mod_boral_Mosquito$row.sigma[[1]]$mean,summary(mod_jSDM_block_Mosquito$mcmc.V_alpha)[[1]]["Mean"], pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)

# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
    jSDM_betas[j,] <- summary(mod_jSDM_block_Mosquito$mcmc.sp[[paste0("sp_",j)]])[[1]][1:np,"Mean"]
  }
boral_betas <- cbind(mod_boral_Mosquito$lv.coefs.mean[,"beta0"],mod_boral_Mosquito$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by Boral", ylab="betas estimated by jSDM", main="Species fixed effect beta_j")
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
    jSDM_lambdas[j,] <- summary(mod_jSDM_block_Mosquito$mcmc.sp[[paste0("sp_",j)]])[[1]][(np+1):(np+nl),"Mean"]
  }
boral_lambdas <- mod_boral_Mosquito$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by Boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda_j")
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- summary(mod_jSDM_block_Mosquito$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"]
}
plot(mod_boral_Mosquito$lv.mean, jSDM_lvs, xlab="latent variables estimated by Boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2")
abline(a=0,b=1,col='red')
# Predictions 
plot(probit_theta_pred_Mosquito, mod_jSDM_block_Mosquito$probit_theta_pred, xlab="probit theta estimated by Boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta")
abline(a=0,b=1,col='red')
```

### Eucalypts dataset

```{r jSDM-boral-Eucalypts}
print(paste(nrow(PA_Eucalypts),"sites and ",ncol(PA_Eucalypts)," species"))
nsp <- ncol(mod_jSDM_block_Eucalypts$model_spec$presences)
nsite <- nrow(mod_jSDM_block_Eucalypts$model_spec$presences)
nl <- mod_jSDM_block_Eucalypts$model_spec$n_latent
np <- nrow(mod_jSDM_block_Eucalypts$model_spec$beta_start)
# Alpha et V_alpha
par(mfrow=(c(1,2)))
plot(mod_boral_Eucalypts$row.coefs[[1]]$mean,summary(mod_jSDM_block_Eucalypts$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by Boral", ylab="alphas estimated by jSDM", main="Site random effect alpha")
abline(a=0,b=1,col='red')
points(mod_boral_Eucalypts$row.sigma[[1]]$mean,summary(mod_jSDM_block_Eucalypts$mcmc.V_alpha)[[1]]["Mean"],
       pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)

# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
    jSDM_betas[j,] <- summary(mod_jSDM_block_Eucalypts$mcmc.sp[[paste0("sp_",j)]])[[1]][1:np,"Mean"]
}
boral_betas <- cbind(mod_boral_Eucalypts$lv.coefs.mean[,"beta0"],mod_boral_Eucalypts$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by Boral", ylab="betas estimated by jSDM", main="Species fixed effect beta_j")
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
    jSDM_lambdas[j,] <- summary(mod_jSDM_block_Eucalypts$mcmc.sp[[paste0("sp_",j)]])[[1]][(np+1):(np+nl),"Mean"]
}
boral_lambdas <- mod_boral_Eucalypts$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by Boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda_j")
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- summary(mod_jSDM_block_Eucalypts$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"]
}
plot(mod_boral_Eucalypts$lv.mean, jSDM_lvs, xlab="latent variables estimated by Boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2")
abline(a=0,b=1,col='red')

# Predictions 
plot(probit_theta_pred_Eucalypts, mod_jSDM_block_Eucalypts$probit_theta_pred, xlab="probit theta estimated by Boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta")
abline(a=0,b=1,col='red')
```

### Frogs dataset

```{r jSDM-boral-frogs}
print(paste(nrow(PA_Frogs),"sites and ",ncol(PA_Frogs)," species"))
nsp <- ncol(mod_jSDM_block_Frogs$model_spec$presences)
nsite <- nrow(mod_jSDM_block_Frogs$model_spec$presences)
nl <- mod_jSDM_block_Frogs$model_spec$n_latent
np <- nrow(mod_jSDM_block_Frogs$model_spec$beta_start)
# Alpha et V_alpha
par(mfrow=c(1,2))
plot(mod_boral_Frogs$row.coefs[[1]]$mean,summary(mod_jSDM_block_Frogs$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by Boral", ylab="alphas estimated by jSDM", main="Site random effect alpha")
abline(a=0,b=1,col='red')
points(mod_boral_Frogs$row.sigma[[1]]$mean,summary(mod_jSDM_block_Frogs$mcmc.V_alpha)[[1]]["Mean"],
       pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)

# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
    jSDM_betas[j,] <- summary(mod_jSDM_block_Frogs$mcmc.sp[[paste0("sp_",j)]])[[1]][1:np,"Mean"]
}
boral_betas <- cbind(mod_boral_Frogs$lv.coefs.mean[,"beta0"],mod_boral_Frogs$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by Boral", ylab="betas estimated by jSDM", main="Species fixed effect beta_j")
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
    jSDM_lambdas[j,] <- summary(mod_jSDM_block_Frogs$mcmc.sp[[paste0("sp_",j)]])[[1]][(np+1):(np+nl),"Mean"]
}
boral_lambdas <- mod_boral_Frogs$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by Boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda_j")
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- summary(mod_jSDM_block_Frogs$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"]
}
plot(mod_boral_Frogs$lv.mean, jSDM_lvs, xlab="latent variables estimated by Boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2")
abline(a=0,b=1,col='red')

# Predictions 
plot(probit_theta_pred_Frogs, mod_jSDM_block_Frogs$probit_theta_pred, xlab="probit theta estimated by Boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta")
abline(a=0,b=1,col='red')
```

### Fungi dataset

```{r jSDM-boral-fungi}
print(paste(nrow(PA_Fungi),"sites and ",ncol(PA_Fungi)," species"))
nsp <- ncol(mod_jSDM_block_Fungi$model_spec$presences)
nsite <- nrow(mod_jSDM_block_Fungi$model_spec$presences)
nl <- mod_jSDM_block_Fungi$model_spec$n_latent
np <- nrow(mod_jSDM_block_Fungi$model_spec$beta_start)
# Alpha et V_alpha
par(mfrow=c(1,2))
plot(mod_boral_Fungi$row.coefs[[1]]$mean,summary(mod_jSDM_block_Fungi$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by Boral", ylab="alphas estimated by jSDM", main="Site random effect alpha")
abline(a=0,b=1,col='red')
points(mod_boral_Fungi$row.sigma[[1]]$mean,summary(mod_jSDM_block_Fungi$mcmc.V_alpha)[[1]]["Mean"],
       pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)

# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
    jSDM_betas[j,] <- summary(mod_jSDM_block_Fungi$mcmc.sp[[paste0("sp_",j)]])[[1]][1:np,"Mean"]
}
boral_betas <- cbind(mod_boral_Fungi$lv.coefs.mean[,"beta0"],mod_boral_Fungi$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by Boral", ylab="betas estimated by jSDM", main="Species fixed effect beta_j")
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
    jSDM_lambdas[j,] <- summary(mod_jSDM_block_Fungi$mcmc.sp[[paste0("sp_",j)]])[[1]][(np+1):(np+nl),"Mean"]
}
boral_lambdas <- mod_boral_Fungi$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by Boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda_j")
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- summary(mod_jSDM_block_Fungi$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"]
}
plot(mod_boral_Fungi$lv.mean, jSDM_lvs, xlab="latent variables estimated by Boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2")
abline(a=0,b=1,col='red')
# Predictions 
plot(probit_theta_pred_Fungi, mod_jSDM_block_Fungi$probit_theta_pred, xlab="probit theta estimated by Boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta")
abline(a=0,b=1,col='red')
```
